---
title: "Acadia National Park Region CBC Results"
author: "Nick Fisichelli, Seth Benz, and Kyle Lima"
date: "4/16/2022"
output:
  html_document:
    theme: readable
  
---

```{r setup, include=FALSE}

library(ggplot2)
library(dplyr)
library(utils)
library(tidyr)
library(sp)
library(ggmap)
library(rgdal)
library(cowplot)
library(sf)

select <- dplyr::select


cbc <- read.csv("cbc_alldata_20220415.csv")
buff1 <- readOGR("mdi_circle.kml")
buff2 <- readOGR("sch_circle.kml")
sp.tab <- read.csv("regional/speciesstats_table_20220415.csv")



#Draw images
cbirds <- ggdraw() +
  draw_image("regional/aa_CumulativeBirds.png")
phours <- ggdraw() +
  draw_image("regional/aa_PartyHours.png")
yso <- ggdraw() +
  draw_image("regional/aa_YearsSpeciesObserved.png")


T2.0 <- cbc %>%
  group_by(CommonName, Year) %>% 
  summarise(Count=mean(Count), PartyHours=mean(PartyHours)) %>% 
  mutate(CountPartyHour = Count/PartyHours)

T2 <- T2.0 %>% 
  group_by(Year) %>%
  summarise(NoSpecies=length(which(Count>0)),
            Birds=sum(Count),
            BirdsPartyHour=sum(CountPartyHour),
            PartyHours=mean(PartyHours))

F1.0 <- T2.0 %>%
  filter(Count > 0)

F1 <- F1.0 %>% 
  group_by(CommonName) %>% 
  summarise(Freq=length(Year),
            TotalBirds=sum(Count),
            TotBirdsPH=sum(CountPartyHour),
            MinYear=min(Year),
            MaxYear=max(Year))

sub.frq <- F1 %>% 
  arrange(desc(Freq)) %>% 
  filter(Freq == 51) %>% 
  select(CommonName, Freq) %>% 
  rename(species=CommonName)

sp.tr <- sp.tab %>% 
  select(species, change)

frgq <- left_join(sub.frq, sp.tr, by = "species")

```

<br>

## Study Area

```{r, fig.height = 8, fig.width = 8, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

##Map
#Create dataframe with circle centers
circpoint <- data.frame(circle  = c("Mount Desert Island", "Schoodic"),
                        latitude = c("44.34", "44.43"),
                        longitude = c("-68.31", "-68.11"))

#Fix non-numeric
circpoint$latitude <- as.numeric(circpoint$latitude)
circpoint$longitude <- as.numeric(circpoint$longitude)

#Fortify for ggplot
buff1f <- fortify(buff1)
buff2f <- fortify(buff2)

sf_buff1 <- st_as_sf(buff1)
sf_buff1_polygons <- st_polygonize(sf_buff1)
shp_buff1 <- as(sf_buff1_polygons, "Spatial")

buff2f$longitude <- buff2f$"long"
buff2f$latitude <- buff2f$"lat"

coordinates(buff2f) <- c("long", "lat")

slot(buff2f, "proj4string") <- slot(shp_buff1, "proj4string")

output <- over(shp_buff1, buff2f, returnList = TRUE) 

output.df <- as.data.frame(output$`0`) #need to remove rows 14:29

buff2.0 <- fortify(buff2)

buff2.0[c(14:29),] <- NA

#Get base map
base.map <- get_stamenmap(
  bbox = c(left = -68.61, bottom = 44.15, right = -67.83, top = 44.6),
  maptype = 'toner-lite',
  zoom = 11)

#Plot
ggmap(base.map) +
  geom_point(data = circpoint, aes(longitude, latitude, fill = circle), 
             shape = 21,
             size = 2.5,
             color = 'black') +
  geom_path(data = buff1f, aes(x=long, y=lat), color = "navyblue") +
  geom_path(data = buff2.0, aes(x=long, y=lat), color = "forestgreen") +
  theme_classic(base_size = 14) +
  ggtitle("Christmas Bird Count (CBC) Circles") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        legend.position = c(0.79, 0.13),
        legend.background = element_rect(fill = "white", color = "black"),
        panel.border = element_rect(color = 'black', size = 1.5, fill = NA)) +
  scale_fill_manual("CBC circle", values = c("navyblue", "forestgreen"))

```
<br>
<br>



## Summary Statistics

We looked at the Schoodic and MDI Christmas Bird Counts from 1971 to 2021. Mean effort between the two circles, recorded as 'party hours', was highest in 1981 with 80.2 party hours, and lowest in 2005 with 25.2 party hours (**Figure 1**). From 1971 to 2021, a collective 162 species were documented on these CBCs. We are stilling seeing new species added each decade, and this doesn't appear to be slowing down. In the 1990s there were 6 new species added, in the first decade of the 2000s there were also 6 new species, and in the most recent decade observers noted 10 new species (**Figure 2**).

**Figure 3** shows that species seen during 5 CBCs or fewer, and species seen on every CBC make up a majority of all species detected on these CBCs. There are `r length(which(F1$Freq==51))` species that have been detected on every CBC since 1971 (**Table 1**). `r length(which(F1$Freq<=5))` species have been recorded on < 5 CBCs (**Table 1**). Of these 28 species that have been detected on every CBC since 1971, `r length(which(frgq$change=="increase"))` species increased, `r length(which(frgq$change=="decrease"))` decreased, and `r length(which(frgq$change=="no change"))` have shown no change over the last 51 years (**Table 2**).

Overall, we found a 53% decline in total count of birds recorded on the CBCs since 1971, and a 43% decline in count of birds per party hour since 1971.



## Summary Figures
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 1**. Number of average party hours by year between the Schoodic and MDI circles. Data provided by the National Audubon Society.'}

plot(phours)

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 2**. Cumulative number of species added each year from the Schoodic and MDI circles. Data provided by the National Audubon Society.'}

plot(cbirds)

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 3**. Number of CBCs that each species has been detected from the Schoodic and MDI circles. Data provided by the National Audubon Society.'}

plot(yso)

```
<br>
<br>

**Table 1**. Species-specific frequency of detection on CBCs, total count since 1971, total count/party hour, and the years in which the highest and lowest count occurred. Data provided by the National Audubon Society.

<br>

<details>
  <summary>**Click to show table**</summary>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

tot.freq <- F1 %>% arrange(desc(Freq))


knitr::kable(tot.freq, col.names = c('Common name', 'Frequency', 'Total count', 'Total count/party hour', 'Min year', 'Max year'))

```
</details>

<br>
<br>


**Table 2**. Species-specific frequency of detection on CBCs, total count since 1971, total count/party hour, and the years in which the highest and lowest count occurred. Data provided by the National Audubon Society.

<br>

<details>
  <summary>**Click to show table**</summary>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

frgq <- frgq %>% 
  arrange(change, species)

knitr::kable(frgq, col.names = c('Common name', 'Frequency', 'Species trend'), align = 'l')

```
</details>

<br>
<br>




## Linear Relationships Through Time

<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 4**. Linear relationship of the number of species by year. Adjusted R-squared:  0.1543, F-statistic: 10.12 on 1 and 49 DF,  p-value: < 0.005. Data provided by the National Audubon Society.'}

T2 %>% 
  ggplot(aes(Year, NoSpecies)) +
  geom_point(shape = 21, size = 1.9, color = "black") +
  geom_smooth(method = "lm", color = "navyblue") +
  theme_bw() +
  labs(title="Number of Species by Year", y="Number of species", x="Year") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0,0), limits = c(50,90)) +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        legend.title = element_blank(),
        axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        strip.text.x = element_text(margin = margin(.2,0,.2,0, "cm"), color = "black", size = "12"), 
        strip.background = element_rect(colour="black", fill="gray"),
        panel.background = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(),
        panel.border = element_rect(color = 'black', fill = NA))

```
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 5**. Linear relationship of the total count of birds by year. Adjusted R-squared:  0.3565, F-statistic: 28.7 on 1 and 49 DF,  p-value: < 0.005. Data provided by the National Audubon Society.'}

T2 %>% 
  ggplot(aes(Year, Birds)) +
  geom_point(shape = 21, size = 1.9, color = "black") +
  geom_smooth(method = "lm", color = "navyblue", na.rm = TRUE) +
  theme_bw() +
  labs(title="Total Count of Birds by Year", y="Tatal count", x="Year") +
  #scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0,0), limits = c(5000,35000)) +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        legend.title = element_blank(),
        axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        strip.text.x = element_text(margin = margin(.2,0,.2,0, "cm"), color = "black", size = "12"), 
        strip.background = element_rect(colour="black", fill="gray"),
        panel.background = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(),
        panel.border = element_rect(color = 'black', fill = NA))

```
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='**Figure 6**. Linear relationship of the total count of birds by year. Adjusted R-squared:  0.02729, F-statistic: 2.403 on 1 and 49 DF,  p-value: 0.1276. Data provided by the National Audubon Society.'}

T2 %>% 
  ggplot(aes(Year, BirdsPartyHour)) +
  geom_point(shape = 21, size = 1.9, color = "black") +
  geom_smooth(method = "lm", colort = "navyblue", na.rm = TRUE) +
  theme_bw() +
  labs(title="Number of Birds/Party Hour by Year", y="Number of birds/party hour", x="Year") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0,0), limits = c(0,550)) +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        legend.title = element_blank(),
        axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        strip.text.x = element_text(margin = margin(.2,0,.2,0, "cm"), color = "black", size = "12"), 
        strip.background = element_rect(colour="black", fill="gray"),
        panel.background = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(),
        panel.border = element_rect(color = 'black', fill = NA))

```
<br>
<br>




## Increasing Species

Top 5 in descending order of severity:

1. Mallard
2. Wild Turkey
3. Bald Eagle
4. Northern Cardinal
5. Harlequin Duck


<br>

## Species showing no change

- Brown Creeper
- Dark-eyed Junco
- American Robin
- Red-breasted Nuthatch
- Song Sparrow

<br>

## Decreasing Species

Top 5 in descending order of severity:

1. Evening Grosbeak
2. American Tree Sparrow
3. Brown-headed Cowbird
4. White-winged Scoter
5. Blue Jay



<br>
<br>



## Example plot
<br>
```{r, fig.height = 7, fig.width = 9, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

ATSP <- cbc %>% 
  filter(CommonName=="American Tree Sparrow") %>% 
  mutate(t.count = Count/60)

ggplot(ATSP, aes(x=Year)) +
    geom_line(aes(y=t.count, color="Count"), size = 1.3) +
  geom_line(aes(y=CountPartyHour, color="Count/party hour"), size = 1.3) +
  scale_y_continuous(name="Count/party hour", sec.axis = sec_axis(~.*60, name="Count")) +
  theme_bw() +
  labs(title="American Tree Sparrow Population Changes Through Time", x="Year") +
  theme(plot.title = element_text(size = "18"),
        axis.text = element_text(color = "black", size = "12"),
        axis.title = element_text(size = "16"),
        axis.line.y.right = element_line(color = "#2166AC", size = 1.2), 
        axis.ticks.y.right = element_line(color = "#2166AC"),
        axis.text.y.right = element_text(color = "#2166AC"), 
        axis.title.y.right = element_text(color = "#2166AC"),
        axis.line.y.left = element_line(color = "#D6604D", size = 1.2), 
        axis.ticks.y.left = element_line(color = "#D6604D"),
        axis.text.y.left = element_text(color = "#D6604D"), 
        axis.title.y.left = element_text(color = "#D6604D"),
        strip.text.x = element_text(margin = margin(.2,0,.2,0, "cm"), color = "black", size = "13"), 
        strip.background = element_rect(colour="black", fill="gray"),
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = 'black', fill = NA, size = 1),
        legend.title = element_text(size=18),
        legend.text = element_text(size=16)) +
  scale_color_manual(name="Statistic",
                     breaks=c("Count/party hour", "Count"),
                     values=c("Count/party hour"="#D6604D", "Count"="#2166AC"))

```
<br>
<br>



## Species Trend Table

The species trend is based soley on the count/party hour estimate and p-value. There are a lot more determinable trends in the count data, but I think we want to use the count/party hour metric to account for effort.

**Table 3** shows that there are `r length(which(sp.tab$change=="increase"))` increasing species, `r length(which(sp.tab$change=="decrease"))` decreasing species, and `r length(which(sp.tab$change=="no change"))` that have shown no change over the last 51 years. The other `r length(which(sp.tab$change=="not enough data"))` species did not have enough data to determine a trend, beacuse they are rarely detected species (vagrants, unusual overwinterers, or irruptive species). 

<br>

**Table 3**. Species trends for all species over the 1971 to 2021 time span. Data provided by the National Audubon Society.

<br>

<details>
  <summary>**Click to show table**</summary>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

sp.tab['species'][sp.tab['species'] == "Gray Jay"] <- 'Canada Jay'

knitr::kable(sp.tab, col.names = c('Species', 'Count estimate', 'Count p-value', 'Count/party hour estimate', 'Count/party hour p-value', 'Species trend'))

```
</details>

<br>
<br>
